What is happening on Twitter? A framework for student research projects with tweets

### 2020-09-15 Rmd/reply-to-reviewers.Rmd

> Comments to the Author
General Comments:

> The revised manuscript “What is happening on Twitter? A framework for student research projects with tweets” provides a case study for facilitating undergraduate research projects using twitter and topic modeling. The manuscript is well-written and is of interest to JSE readers, and is a strong “datasets and stories” article.

> I have only two general suggestions: first, it would be nice to have a brief description of the two undergraduate projects within the manuscript. While I understand the data is no longer available, and think it was the right decision to focus on the case study for methods and results, including a short paragraph about each of the undergraduate projects you supervised would help to give the reader a sense of the range of possibilities of a twitter-based project. This is briefly mentioned at the end of the manuscript but it would help to see the motivations for such a project earlier.

**Thank you for this suggestion! We've now added text that describes the two projects early in the manuscript.**

@@

> Second, while I appreciate the concise and clear writing, adding transitions between the sections would help the reader anticipate what is coming. As one example, prior to Sec 5.2, there is no indication that the authors will walk through (a) querying twitter API (b) accessing full tweets via ID numbers (c) parsing JSON.

**We appreciate this suggestion, too. We've worked to add transitions between sections with the goal of better guiding the reader's expectations.**

@@

**We've now added a statement like this to the manuscript to reflect this point.**

> - Page 4 Sec 4.2: I agree with the authors that differences in student research interests and goals should be an important consideration and it would be nice to see some examples.

**We've added two examples from our students.**

> - Page 4 Line 82: Would this type of project also be appropriate for a summer research experience?

**Yes, we've amended the text to reflect this.**

> - Page 4 Line 88-92: This wording could be a little more clear: is the case study essentially a replicate of a student project? It would be helpful to see an overview of what the original project was, and then introduce the case study as a replicate.

**We've reworded this to clarify the point. Yes, the case study essentially replicates a student project.**

> - Page 5 Sec 5.2: would this process theoretically work for both windows and Mac machines? Is it necessary to use a crontab-like process, or could it be done entirely within R?

@@

> - Page 8 Figure 1:“beta” has not been mentioned before appearing in this graph (related to previous comment)

**You make a good point. We think that the equations that we've added (to address the previous comment) also resolve this.**

> - Page 11 Line 162-163: “… our interest is in the transient appearance of a new topic” — this could be made more clear at the beginning of the case study. Another sentence or two motivating what to expect for results would be helpful.

**We agree with this suggestion, too. We have tried to clarify it by adding a sentence to this section.**

@@

> I think sections 3 and 4 could be combined into one section that is perhaps “Structure of mentored research” or something like that. Perhaps there would be fewer subsections under that section header, but more details on how the research was structured.

**We agree. We've removed many of the subsection headings and used your suggested section title.**

> Throughout the existing sections, the verb tenses were strange. Since this is now a case study, I expected everything to be grounded in what you actually did. But, on page 3 it says, “These are our four learning objectives.” Why not, “were”? Did you develop the learning objectives after the fact?

**We didn't write out learning objectives at the time of working with these students, but we did informally discuss them. Back then (2015), we were new to research mentoring. We now are more explicit while designing projects with students.**

@@

> Section 4.2 says that “an initial brainstorming session may clarify.” Why not “helped”? Is an initial brainstorming session just something you recommend/plan to do in the future, or did you actually do it with those students? You say that you anticipate that sharing completed student project reports will guide future students, but this is from the 2015-2016 school year. Since this project, has it become easier to talk to students about goals, grounded in examples like this? Back to the students in 2015-2016, what goals did you eventually land on?

**Thank you for raising this issue. We agree that it needs clarification. We've added text to address this at lines.**

> I assume that those discussions with students involved conversations about the learning objectives from the previous section, as well as the expectations for students (presentation at the end of the year, perhaps some amount of weekly work?).

**Yes, although we were new research mentors when we began working with these two students. Regrettably, we didn't structure our conversations as we would now.**

@@

> Reviewer: 3

> Comments to the Author
The authors describe a framework for student research using data available using the twitter API.  This is an interesting and relevant paper since many other uses of Twitter data have proven to be somewhat superficial in terms of the richness of data.  The paper is grounded in research on projects.

> My main issue relates to the question being answered here: what actionable insights are being extracted from these data?  Is it really just that Memorial Day words show up on Memorial Day?  Are there other questions that could be considered (even if the authors don’t answer them)?  I’m left with the concern that Twitter data is cumbersome to deal with (in terms of scale and restrictions on use of the API) and not particularly information rich.

**Thank you. We've attempted to address these concerns by writing more about the original two student projects.**

### 2020-09-15 Rmd/jse-special-issue-2020-october.Rmd
deleted blank line to combine paragraphs
added text on NASEM report and data acumen.

@@

@@

Our two students had diverse interests, and, initially, they had little experience in 
articulating research goals. We engaged each in a brainstorming session to 
clarify their interests and encourage them to think critically about research goals under 
the time constraints of their academic schedules. 
Additionally, we anticipate that sharing completed student project reports with 
future trainees will guide thinking about the scope of possible projects (Supplementary files).

We briefly describe the two student projects to give readers a better sense 
of research possibilities with tweets. Our first student examined relationships over time between 
stock market index prices and tweet sentiment. For each day in her 12-month study period, 
she identified stock market-related tweets with a key word search. With the complete
texts of stock market-related tweets for each day, she calculated a daily sentiment score and plotted it over time. 
Her sentiment score reflected presence of emotion-associated terms 
(*eg*., "happy", "sad", "mad", "scared") 
in tweet texts. Days with more net positive emotion words in the collected tweets received a higher 
(positive) daily sentiment score, while days with more net negative words received a negative daily 
sentiment score. 
For her final project, she presented plots over time of her daily sentiment scores and daily closing 
prices of the Standard and Poor's 500 index. She also explored time series analysis methods to quantify 
relationships between index prices and sentiment scores.

@@

To illustrate the value of Twitter data, we present below a 
reproducible case study. It is essentially a reproduction of our second student's project, but at a 
distinct time period. In it, we aim to detect a social media 
event by examining topic content over time. We use Latent Dirichlet 
Allocation (LDA) models [@blei2003latent] to infer topics on three consecutive days centered 
on Memorial Day 2020. We chose this example case study, 
instead of the student projects, because of limited data 
availability for the student projects. Despite this, the case study 
illustrates the strategy and methods for one of the student 
projects.

@@

The goal for LDA is to infer both the distribution over topics and 
the topics [@blei2003latent]. A topic, in this setting, is a distribution over 
the vocabulary (the collection of all words in a corpus).

Inference for latent Dirichlet allocation models is performed by either 
sampling from the posterior distribution or through variational methods. Researchers
have devised a variety of Gibbs sampling techniques for these models [@porteous2008fast]. 
Variational methods, while using approximations to the posterior distribution, offer the 
advantage of computational speed [@blei2017variational]. We used variational methods below in our case study.

@@

Assigning meaning to topics is an active research area [@chang2009reading]. Since
our interest is in the transient appearance of a new topic, we don't attempt to
assign meaning to every topic in our models. We see that topic 7 from May 25 has several 
words that suggest Memorial Day: memorial, remember, honor, country. A similar topic is not 
seen on May 24 or May 26. Some topics persist, with distinct word probabilities, 
across the three days. For example, we see that President Trump features prominently in all 
three models. On May 26, topic 10 reflects discussion of the Amy 
Cooper Central Park incident (https://www.nytimes.com/2020/05/26/nyregion/amy-cooper-dog-central-park.html).

The murder of George Floyd occurred on May 25. Our last 
examined time period, from 12:00 pm to 12:05 pm (Eastern USA time zone) on May 26, occurred after Floyd's murder, 
yet we didn't detect this event in our ten-topic LDA model. Several considerations 
may account for this. While outrage at the murder eventually spread worldwide, there may have been few 
Floyd-related tweets during our collection time on May 26, less than 24 hours after the murder and video release. 
Had we extended our analysis to May 27 and beyond, we may have identified George Floyd-related topics.

>>>>

In particular, our project skills reflect elements of "data acumen", as defined 
in a report from the U.S.A. National Academies of Sciences, Engineering, and Medicine (NASEM) [@national2018data]. For example, our skills "Acquire data from internet sources" and 
"Develop data science strategies to address a research question" implicitly require 
a trainee, in the language of the NASEM report [@national2018data], to "ingest, clean, and then 
wrangle data into reliable and useful forms" and to "combine many existing programs or codes
into a workflow that will accomplish some important task". Additionally, we tailored our list 
of project skills with the assumption that students would work in R. R use is not required for such projects, but it is a 
convenience for many of our students.

@@

Having decided on four learning objectives for students, we next established
mentoring relationships with each student.

@@

Our two students had diverse interests, and, initially, they had little experience in 
articulating research goals. We engaged each in a brainstorming session to 
clarify their interests and encourage them to think critically about research goals under 
the time constraints of their academic schedules. 
We briefly describe the two student projects to give readers a better sense 
of research possibilities with tweets.

Our first student examined relationships over time between 
stock market index prices and tweet sentiment. For each day in her 12-month study period, 
she identified stock market-related tweets with a key word search. With the complete
texts of stock market-related tweets for each day, she calculated a daily sentiment score and plotted it over time. 
Her sentiment score reflected presence of emotion-associated terms 
(*eg*., "happy", "sad", "mad", "scared") 
in tweet texts. Days with more net positive emotion words in the collected tweets received a higher 
(positive) daily sentiment score, while days with more net negative words received a negative daily 
sentiment score. 
For her final project, she presented plots over time of her daily sentiment scores and daily closing 
prices of the Standard and Poor's 500 index. She also explored time series analysis methods to quantify 
relationships between index prices and sentiment scores.

@@

To illustrate the value of Twitter data, we present below a 
reproducible case study. It is essentially a reproduction of our second student's project, but at a 
distinct time period. In it, we aim to detect a social media 
event by examining topic content over time. We use Latent Dirichlet 
Allocation (LDA) models [@blei2003latent] to infer topics on three consecutive days centered 
on Memorial Day 2020. We chose this example case study, 
instead of the student projects, because of limited data 
availability for the student projects. Despite this, the case study 
illustrates the strategy and methods for one student 
project.

We next discuss case study design, tweet collection, and tweet structure, before turning 
to statistical methods for the case study.

@@

The goal for LDA is to infer both the distribution over topics and 
the topics [@blei2003latent]. A topic, in this setting, is a distribution over 
the vocabulary (the collection of all words in a corpus). 
Inference for latent Dirichlet allocation models is performed by either 
sampling from the posterior distribution or through variational methods. Researchers
have devised a variety of Gibbs sampling techniques for these models [@porteous2008fast]. 
Variational methods, while using approximations to the posterior distribution, offer the 
advantage of computational speed [@blei2017variational]. We used variational methods below in our case study.

@@

Assigning meaning to topics is an active research area [@chang2009reading]. Since
our interest is in the transient appearance of a new topic, we don't attempt to
assign meaning to every topic in our models. Instead, we anticipate that discussions on Twitter are a mixture of topics that endure over weeks or months and subjects that appear and disappear quickly. We see that topic 7 from May 25 has several 
words that suggest Memorial Day: memorial, remember, honor, country. A similar topic is not 
seen on May 24 or May 26. Some topics persist, with distinct word probabilities, 
across the three days. For example, we see that President Trump features prominently in all 
three models' results. On May 26, topic 10 reflects discussion of the Amy 
Cooper Central Park incident (https://www.nytimes.com/2020/05/26/nyregion/amy-cooper-dog-central-park.html).

The murder of George Floyd occurred on May 25, 2020. Our last 
examined time period, from 12:00 pm to 12:05 pm (Eastern USA time zone) on May 26, occurred after Floyd's murder, 
yet we didn't detect this event in our ten-topic LDA model. Several considerations 
may account for this. While outrage at the murder eventually spread worldwide, there may have been few 
Floyd-related tweets during our collection time on May 26, less than 24 hours after the murder and video release. 
Had we extended our analysis to May 27 and beyond, we may have identified George Floyd-related topics.
